msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-06-07 paper.tex
Update mutation models.

\begin{abstract}
Coalescent simulation is a key tool in population genetics and
and has been integral to coalescent theory since its earliest days.
Because of the ease at which the basic model can be simulated,
a large number of different simulators have been developed. However,
the coalescent with recombination is far more challenging to simulate,
and, until recently, it was not possible to simulate efficiently.
The \msprime\ software has revolutionised
coalescent simulation, making it possible to simulate millions
of whole genomes for the first time. We summarise the many features
of \msprime\ 1.0, which is built around a core model of efficiently
implementing recombination via the succinct tree sequence data
structure. We advocate a community oriented open-source development
process as a way to reduce duplication of effort and increase
the quality of simulation software.
\end{abstract}

@@

The coalescent
process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
models the ancestry of set of sampled genomes. Given some sampled
genomes, the coalescent provides a mathematical description of the
genealogical tree relating these to each other. Because of its
elegance and simplicity, the coalescent is now central to much of
population genetics and
genomics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
Simulation has been a vital
tool in coalescent theory since its earliest days~\citep{hudson1983testing},
and dozens of different simulation tools have been
developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Two developments of recent years, however, have made the majority
of these simulators of little relevance to modern datasets.
% Some citations here? Darwin tree of life maybe?
Firstly, whole genome sequencing technology---and in particular fourth-generation
sequencing technologies---have made chromosome-level assemblies
commonplace. Thus, regarding input data as a series of unlinked loci
(a previously reasonable assumption), is no longer defensible and
recombination must be fully accounted for. While the
single-locus coalescent can be simulated in constant
time~\citep{hudson1990gene}, the coalescent with recombination is
far more challenging to simulate.
% Sentence is too long...
Programs such as Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination,
and, until recently, we had to resort to
approximate models of recombination~\citep{mcvean2005approximating,staab2014scrm}
if we wished to simulate whole chromosomes.

The second development that has made the majority of
present day simulators inapplicable to modern data sets is the
exponential increase in sample size. Human datasets such
UK Biobank~\citep{bycroft2018genome} and
gnomAD~\citep{karczewski2019variation} now contain hundreds of
thousands of genomes, and there is every reason to believe that
datasets of millions will soon be commonplace~\citep{stephens2015big}.
Classical simulators such as \ms\ and even fast approximate simulators
such as \scrm\ simply cannot cope with this scale. Even if we
spend months of CPU time on running simulations with millions of samples,
the output would require terabytes of storage space to store and
days of CPU time to parse~\citep{kelleher2016efficient}.

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has profoundly changed
what is possible. Through the use of efficient data structures, it is
now possible to simulate millions of whole chromosomes, without resorting
to approximations. The ``succinct tree
sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring},
introduced as part of \msprime\, makes it possible to store simulations
of millions of whole chromosomes in a few gigabytes, several orders
of magnitude smaller than the standard
Newick~\citep{felsenstein1989phylip} formats. This data structure has
also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring}
and calculation of population genetic statistics~\citep{ralph2019efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making a highly efficient and flexible platform for population
genetic simulation.
% This is a bit flat --- anything else to say at the end here?
This paper marks the release of \msprime\ 1.0.

@@

We also support variable recombination rates along a chromosome.
This allows us to model both recombination hotspots like
\texttt{msHOT}~\citep{hellenthal2007mshot} and also empirical
recombination maps like those supported by X Y Z.
(What about ~\cite{wang2014new}?)
We have also implemented gene
conversion following the model defined by~\cite{wiuf2000coalescent}
and used in \ms.
% % Uncomment this sentence, when the corresponding issue in the msprime github repo is resolved.
% Furthermore, msprime is now able to simulate gene conversion events that are
% initiated proportional to the local recombination rate specified by the recombination map.

@@

@@

\subsection*{Seedbank coalescent}
Seedbank coalescent. [If we implement this fill in please Thibaud]

@@

Thus, development of \msprime\ is not completed and much remains
to be done. For example, one key aim of the 1.0 series is to
improve the mutation models supported which are limited to
simple infinite-sites currently. The ability to simulate
sophisticated mutation models like Seq-Gen~\citep{rambaut1997seq},
Dawg~\citep{cartwright2005dna} and Pyvolve~\citep{spielman2015pyvolve} at the megasample scale would be
useful in many applications. Another key focus of development
will be [WHAT?]

>>>>

\begin{abstract}
Coalescent simulation is a key tool in population genetics and
has been integral to coalescent theory since its beginnings.
Because of the ease at which the basic model can be simulated,
a large number of different simulators have been developed. However,
the coalescent with recombination is far more challenging to simulate,
and, until recently, it was not possible to simulate efficiently.
The \msprime\ software has revolutionised
coalescent simulation, making it possible to simulate millions
of whole genomes for the first time. We summarise the many features
of \msprime\ 1.0, which is built around a core model of efficiently
implementing recombination via the succinct tree sequence data
structure. We advocate a community oriented open-source development
process as a way to reduce duplication of effort and increase
the quality of simulation software.
\end{abstract}

@@

The coalescent
process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
models the ancestry of set of sampled genomes. Given some sampled
genomes, the coalescent provides a mathematical description of the
genealogical tree relating these to each other. Because of its
elegance and simplicity, the coalescent is now central to much of
population genetics and
genomics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
Simulation has been a vital
tool in coalescent theory since its earliest days~\citep{hudson1983testing},
and dozens of different simulation tools have been
developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Two developments of recent years, however, have made the majority
of these simulators of little relevance to modern datasets.
% Some citations here? Darwin tree of life maybe?
Firstly, whole genome sequencing technology---and in particular fourth-generation
sequencing technologies---have made chromosome-level assemblies
commonplace. Thus, regarding input data as a series of unlinked loci
(a previously reasonable assumption), is no longer defensible and
recombination must be fully accounted for. While the
single-locus coalescent for $n$ samples can be simulated in $O(n)$
time~\citep{hudson1990gene}, the coalescent with recombination is
far more challenging.
% Sentence is too long...
Programs such as Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination,
and, until recently, approximate models of
recombination~\citep{mcvean2005approximating,staab2014scrm}
were necessary to simulate whole chromosomes.

The second development that has made the majority of
present day simulators inapplicable to modern data sets is the
exponential increase in sample size. Human datasets such
UK Biobank~\citep{bycroft2018genome} and
gnomAD~\citep{karczewski2019variation} now contain hundreds of
thousands of genomes, and there is every reason to believe that
researchers will be soon be routinely working with
datasets containing millions of samples~\citep{stephens2015big}.
Classical simulators such as \ms\ and even fast approximate simulators
such as \scrm\ simply cannot cope with this scale. Even if we
spend months of CPU time on running simulations with millions of samples,
the output would require terabytes of storage space to store and
days of CPU time to parse~\citep{kelleher2016efficient}.

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has profoundly changed
what is possible. Through the use of efficient data structures, it is
now possible to simulate millions of whole chromosomes, without resorting
to approximations. The ``succinct tree
sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
introduced as part of \msprime\, makes it possible to store simulations
of millions of whole chromosomes in a few gigabytes, several orders
of magnitude smaller than the standard
Newick~\citep{felsenstein1989phylip} based formats. This data structure has
also lead to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic statistics~\citep{ralph2019efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making a highly efficient and flexible platform for population
genetic simulation.
% This is a bit flat --- anything else to say at the end here?
This paper marks the release of \msprime\ 1.0.

@@

We also support variable recombination rates along a chromosome.
This allows us to model both recombination hotspots like
\texttt{msHOT}~\citep{hellenthal2007mshot} and also empirical
recombination maps like those supported \texttt{cosi2}
\citep{shlyakhter2014cosi2}
(What about ~\cite{wang2014new}?)
We have also implemented gene
conversion following the model defined by~\cite{wiuf2000coalescent}
and used in \ms.
% % Uncomment this sentence, when the corresponding issue in the msprime github repo is resolved.
% Furthermore, msprime is now able to simulate gene conversion events that are
% initiated proportional to the local recombination rate specified by the recombination map.

@@

\textbf{TODO:} It's important that we acknowledge that (a) the coalescent
with recombination does become quite slow for large $N_e$; (b) our
implementation of the SMC is naive; and (c) for sufficiently large
$N_e$ programs like \texttt{scrm} will be faster.

@@

\subsection*{Finite sites mutations}
Version 0.x of \texttt{msprime} provided a very simple mutation model,
supporting only the infinite sites model.
With version 1.0, we now support a large number of mutation models,
including
JC69~\citep{jukes1969evolution},
F84~\citep{felsenstein1996hidden},
and
GTR~\citep{tavare1986some} nucleotide models
and
the BLOSUM62~\citep{henikoff1992amino}
and PAM~\citep{dayhoff1978} amino acid models.
The general model accepts an arbitrary transition matrix between
any number of alleles (which can be arbitrary unicode strings),
and so other models such as the Kimura two and three
parameter models~\citep{kimura1980simple,kimura1981estimation}
parameters models can be easily
and efficiently defined in user code.
Full support for the original infinite sites model is also maintained.

\textbf{TODO} Compare performance and features with
Seq-Gen~\citep{rambaut1997seq},
Dawg~\citep{cartwright2005dna} and Pyvolve~\citep{spielman2015pyvolve}

@@

Thus, development of \msprime\ is not completed and much remains
to be done.

% We need to acknowledge that for long genomes and for large population
% sizes msprime's performance does drop off and SMC simulators like
% scrm become more efficient.
Msprime's performance for large sample sizes is excellent, but
simulating species with large population sizes is still challenging.
An important avenue for future development is to improve performance
in these scenarios by incorporating data structures such as skip lists
and implementing more efficient coalescent approximations, as done
in \texttt{cosi2}~\citep{shlyakhter2014cosi2}.
